Algebraic Principles for Concurrency Verification Tools

### 2013-11-11 Paper/paper.tex
Updates to paper

\DeclareFontFamily{U}{wncy}{}
\DeclareFontShape{U}{wncy}{m}{n}{<->wncyss10}{}
\DeclareSymbolFont{mcy}{U}{wncy}{m}{n}
\DeclareMathSymbol{\sh}{\mathord}{mcy}{"78} 
\DeclareMathOperator{\sha}{\sh}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\fin}{finite}
\DeclareMathOperator{\rely}{rely}
\DeclareMathOperator{\guar}{guar}
\DeclareMathOperator{\letters}{letters}
\DeclareMathOperator{\wlp}{wlp}
\DeclareMathOperator{\test}{test}
\DeclareMathOperator{\eval}{eval}

@@

The exchange law is initially appealing as it provides a way to derive
many properties of parallel composition from a single law. For example
one can show that $\|$ must be commutative, or that $x\cdot y \le
x\|y$. However, there are many interesting models which do not satisfy
this law. First, one could consider fair parallel composition
$x\parallel_f y$ in a model with infinite, or non-terminating
programs. In this model, $x \cdot y \not\leq x \parallel_f y$ in the
case where $x$ is non-terminating. A second example would be
predicate-transformer style models for separation logic, where $\|$ is
interpreted as separating conjunction. In such models, $\|$ and
$\cdot$ have different units, which, in the presence of the exchange
law implies that $\cdot$ is equal to $\|$.

@@

For programs, the star represents finite iteration of a program $x$
zero or more times. For iteration where $x$ must be executed at least
once, we write $x^+$. One can similarly axiomatise a parallel star
operation, but in practice this seems to be a less useful operation,
although one could see it as an operator that spawns an arbitrary but
finite number of processes.

@@

One might be tempted to add a test subalgbra to concurrent Kleene
algebra in the style of KAT. In such an algebra, can the Kozen-style
encoding of hoare triples above can be used to derive Owicki-Gries
style concurrency rules? The answer seems to be no---an algebraic
approach imposes strong compositionality requirements (a point we will
return to later), so it should come as no suprise that a
non-compositional method such as Owicki-Gries cannot be so easily be
done algebraically.

@@

TODO: Define algebra w/ c.

@@

The rely-guarantee elements in this model are sets containing all the
words which can be built from some set of state pairs in
$\Sigma\times\Sigma$. We define a function $\langle R\rangle$ which
lifts a relation $R$ to a language containing length 1 words for each
pair in $R$. The set of rely-guaratee elements RG is then defined as
$\{r.\; \exists R. r = \langle R\rangle^*\}$. We can show that
$(\mathcal{P}((\Sigma\times\Sigma)^*), RG, \cup, \cdot,\sha,
\emptyset, \{\epsilon\}, C)$ is a rely-guarantee algebra.

@@

  \path [rel] (s11) -- (s22);
  \path [rel] (s12) -- (s22);
  \path [rel] (s23) -- (s32);
\end{tikzpicture}
\end{center}
The language contains the following six words
\begin{align*}
(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_1,\sigma_2)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3),\\
(\sigma_2,\sigma_2)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_1,\sigma_1)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3),\\
(\sigma_1,\sigma_2)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_2,\sigma_2)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3),
\end{align*}
but only the first,
$(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is
consistent. This word is highlighted with solid arrows in the
diagram above. Now imagine we place the single state pair $(\sigma_2,
\sigma_3)$ into the above language, we might end up with a language as
below:

@@

By performing this shuffle action, we no longer have a consistent word
from $\sigma_1$ to $\sigma_3$, but instead a consistent word from
$\sigma_2$ to $\sigma_3$ and $\sigma_1$ to $\sigma_3$. The new
consistent word was constructed by making a previously inconsistent
word consistent by `filling in the gaps' in that word. This shows that
the shuffle operator can take two inconsistent words and shuffle them
together in such a way that it generates many consistent words. If we
only considered consistent words, we would be unable to define such a
shuffle operator, and would instead have to rely on some operational
semantics to generate traces instead, which would be less
algebraic. This also clearly demonstrates why shuffle is not well
defined on the subalgebra we get by intersecting with $C$.

\newpage

Following~\cite{brookes_full_1993} and~\cite{dingel_refinement_2002},
we inductively generate the stutter/mumble language $w^\dagger$ for a
word $w$ in a language over $\Sigma^2$ as follows: Assume $R,S,T \in
\Sigma^2$ and $u,v,w \in (\Sigma^2)^*$. First, $Iw \in
w^\dagger$. Secondly, if $uRv \in x^\dagger$ then $uIRv \in
x^\dagger$ and $uRIv \in x^\dagger$ (\emph{stuttering}). Thirdly, if
$uRSv \in w^\dagger$ then $u(R;S)v \in w^\dagger$
(\emph{mumbling}). The stutter/mumble closure of a language is
thus defined by lifting this operation to languages.
\begin{align*}
X^\dagger = \bigcup\{x^\dagger.\; x \in X\}
\end{align*}

@@

In the trace model, assignment is defined as
\begin{align*}
  x := e = \bigsqcup v.\,\test\{\sigma.\,\eval(\sigma,e) = v\} \cdot x \leftarrow v
\end{align*}
where $x \leftarrow v$ denotes the atomic command which assigns the value $v$ to $x$.

>>>>

\DeclareFontFamily{U}{wncy}{}
\DeclareFontShape{U}{wncy}{m}{n}{<->wncyss10}{}
\DeclareSymbolFont{mcy}{U}{wncy}{m}{n}
\DeclareMathSymbol{\sh}{\mathord}{mcy}{"78} 
\DeclareMathOperator{\sha}{\sh}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\fin}{finite}
\DeclareMathOperator{\rely}{rely}
\DeclareMathOperator{\guar}{guar}
\DeclareMathOperator{\letters}{letters}
\DeclareMathOperator{\wlp}{wlp}
\DeclareMathOperator{\test}{test}
\DeclareMathOperator{\eval}{eval}
\DeclareMathOperator{\Id}{Id}
\DeclareMathOperator{\edn}{end}

@@

The exchange law is initially appealing as it provides a way to derive
many properties of parallel composition from a single law. For example
one can show that $\|$ must be commutative, or that $x\cdot y \le
x\|y$. However, there are many interesting models which do not satisfy
this law. First, one could consider fair parallel composition
$x\parallel_f y$ in a model with infinite, or non-terminating
programs. In this model, $x \cdot y \not\leq x \parallel_f y$ in the
case where $x$ is non-terminating. A second example would be
predicate-transformer style models for separation logic, where $\|$ is
interpreted as separating conjunction. In such models, $\|$ and
$\cdot$ have different units, which, in the presence of the exchange
law, implies that $\cdot$ is equal to $\|$.

@@

For programs, the star represents finite iteration of a program $x$
zero or more times. For iteration where $x$ must be executed at least
once, we write $x^+$. One can similarly axiomatise a parallel star
operation.

@@

@@

As such we redefine our rely-guarantee algebra from section
\ref{sec:RG} as a structure $(K,RG,+,\sqcap,\cdot,\|,^\star,0,1,c)$
which, in addition to the rules in section \ref{sec:RG} satisfies both
(\ref{con1}) and (\ref{con2}).

@@

The rely-guarantee elements in this model are sets containing all the
words which can be built from some set of state pairs in
$\Sigma\times\Sigma$. We define a function $\langle R\rangle$ which
lifts a relation $R$ to a language containing words of length one for
each pair in $R$. The set of rely-guaratee elements RG is then defined
as $\{r.\; \exists R. r = \langle R\rangle^*\}$. We can show that
$(\mathcal{P}((\Sigma\times\Sigma)^*), RG, \cup, \cdot,\sha,
\emptyset, \{\epsilon\}, C)$ is a rely-guarantee algebra.

Since $\langle R\rangle$ is atomic, it satisfies several useful properties, such as,
\begin{align*}
\langle R\rangle^\star \| \langle S\rangle &= \langle R\rangle^\star; \langle S\rangle; \langle R\rangle^\star\\
\langle R\rangle^\star \| \langle S\rangle^\star &= (\langle R\rangle^\star; \langle S\rangle^\star)^\star
\end{align*}

@@

  \path [rel] (s11) -- (s22);
  \path [rel] (s12) -- (s22);
  \path [rel] (s23) -- (s32);
\end{tikzpicture}
\end{center}
The language contains the following six words
\begin{align*}
(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_1,\sigma_2)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3),\\
(\sigma_2,\sigma_2)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_1,\sigma_1)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3),\\
(\sigma_1,\sigma_2)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_2,\sigma_2)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3),
\end{align*}
where only the first,
$(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is
consistent. This word is highlighted with solid arrows in the
diagram above. Now if we shuffle the single state pair $(\sigma_2,
\sigma_3)$ into the above language, we might end up with a language as
below:

@@

By performing this shuffle action, we no longer have a consistent word
from $\sigma_1$ to $\sigma_3$, but instead a consistent word from
$\sigma_2$ to $\sigma_3$ and $\sigma_1$ to $\sigma_3$. These new
consistent words were constructed from previously inconsistent
words---the shuffle operator can take two inconsistent words and
shuffle them together in such a way to generate many consistent
words. If we only considered consistent words, \'a la Aczel traces, we
would be unable to define such a shuffle operator directly on the
traces themselves, and would instead have to rely on some operational
semantics to generate traces.

Following~\cite{brookes_full_1993} and~\cite{dingel_refinement_2002},
we inductively generate the mumble language $w^\dagger$ for a word $w$
in a language over $\Sigma^2$ as follows: Assume $R,S,T \in \Sigma^2$
and $u,v,w \in (\Sigma^2)^*$. First, $w \in w^\dagger$. Secondly, if
$uRSv \in w^\dagger$ then $u(R;S)v \in w^\dagger$. This operation is
lifted to languages in the obvious way as
\begin{align*}
X^\dagger = \bigcup\{x^\dagger.\; x \in X\}.
\end{align*}
Stuttering is represented as a rely condition $\langle \Id\rangle^\star$
where $\Id$ is the identity relation. Two languages $X$ and $Y$ are
equal under stuttering if $\langle \Id\rangle^\star \| X =_\pi \langle
\Id\rangle^\star \| Y$.

Stuttering and mumbling are important for defining \emph{tests}. A
test is any language $P$ where $P \le \langle \Id\rangle$. We
write $\test(P)$ for $\langle \Id_P\rangle$. Without any
stuttering and mumbling the traces $\test(P); \test(Q)$ and
$\test(P \cap Q)$ are incomparable, as all words in the former have
length two, while all the words in the latter have length one. With mumbling we have that
\begin{align*}
\test(P \cap Q) \le_\pi \test(P); \test(Q)
\end{align*}
as the longer words in $\test(P);\test(Q)$ can be mumbled down into
the shorter words of $\test(P\cap Q)$, whereas with stuttering we
have the opposite,
\begin{align*}
\langle I\rangle^\star \| (\test(P);\test(Q)) \le_\pi \langle I\rangle^\star \| \test(P \cap Q).
\end{align*}
For the rest of the paper, we assume that all languages are
implicitly mumble closed.

Using tests, we can encoding common program statements such as if
statements and while loops as in Kozen's Kleene algebra with tests.
\begin{align*}
\text{if } P \text{ \{ } X \text { \} else \{ } Y \text{ \}} &= \test(P); X + \test(- P); Y\\
\text{while } P \text{ \{ } X \text{ \}} &= (\test(P);X)^\star;\test(- P)
\end{align*}
Finally, we define the operator $\edn(P)$ which is contains all the
words which end in a state satisfying $P$. Some useful properties of $\edn$ include
\begin{align*}
\edn(P); \test{Q} &\le_\pi \edn(P \cap Q),\\
\test(P) &\le \edn(Q),\\
\text{range}(\Id_P \circ R) \le P \implies \edn(P); \langle R\rangle^\star &\le_\pi \edn(P).
\end{align*}

In the trace model, assignment is defined as
\begin{align*}
  x := e = \bigcup v.\,\test\{\sigma.\,\eval(\sigma,e) = v\} \cdot x \leftarrow v
\end{align*}
where $x \leftarrow v$ denotes the atomic command which assigns the
value $v$ to $x$. The $\eval$ function evaluates an expression $e$ in
the state $\sigma$. Using this definition of assignment we can derive
the assignment rule
\begin{align*}
&\text{unchanged}(\text{vars}(e)) \cap \text{preserves}(P) \cap \text{preserves}(P[x/e]),\\
&\text{unchanged}(- {x})\\
&\vdash \{\edn(P)\} x := e \{\edn(P[x/e])\}.
\end{align*}

The rely condition states that the environment is not allowed to
modify any of the variables used when evaluating $e$, i.e. those
variables must remain unchanged, and also that the environment must
preserve the precondition and postcondition of the assignment
statement. In turn, the assignment statement itself guarantees that it
leaves every variable other than $x$ unchanged. Preserves and unchanged are defined as
\begin{align*}
\text{preserves}(P) &= \langle \{(\sigma,\sigma').\; P(\sigma) \implies P(\sigma')\}\rangle^\star,\\
\text{unchanged}(X) &= \langle \{(\sigma,\sigma').\; \forall v\in X.\; \sigma(v) = \sigma'(v)\}\rangle^\star.
\end{align*}
We also defined two futher rely conditions, increasing and decreasing,
which are defined much like unchanged except they only require that variables
increase or decrease, rather than stay the same.

@@

Figure \ref{fig:findp} shows the classic FINDP program, which has been
used by numerous authors including Owicki~\cite{owicki},
Jones~\cite{jones}, de Roever~\cite{derover} and
Hayes~\cite{hayes}. The program finds the least element of an array
satisfying the predicate $P$. The index of the first element
satisfying $p$ is placed in the variable $f$. If there are no elements
of the array satisfying $P$ then $f$ will be set to the length of the
array. The program has two subprograms, $A$ and $B$, which run in
parallel, one of which searches the even indices while the other
searches the odd indices. A speedup over a sequential implementation
is acheived as $A$ will terminate when $B$ finds an element of the
array satisfying $P$ which is less than $i_A$.

\begin{figure}
\[
\begin{aligned}
&f_A := \text{len}(\text{array});\\
&f_B := \text{len}(\text{array});\\
&\left(
\begin{aligned}
&i_A = 0\\
&\text{while } i_A < f_A \land i_A < f_B \text{ \{ }\\
&\qquad\text{if } P(\text{array}[i_A]) \text{ \{}\\
&\qquad\qquad f_A := i_A\\
&\qquad\text{\} else \{}\\
&\qquad\qquad i_A := i_A + 2\\
&\qquad\text{\}}\\
&\text{\}}
\end{aligned}
\middle\|
\begin{aligned}
&i_B = 1\\
&\text{while } i_B < f_A \land i_B < f_B \text{ \{}\\
&\qquad\text{if } P(\text{array}[i_B]) \text{ \{}\\
&\qquad\qquad f_B := i_B\\
&\qquad\text{\} else \{}\\
&\qquad\qquad i_B := i_B + 2\\
&\qquad\text{\}}\\
&\text{\}}
\end{aligned}
\right);\\
&f = \text{min}(f_A,f_B)
\end{aligned}
\]
\caption{FINDP Program}
\label{fig:findp}
\end{figure}

To prove the correctness of FINDP, we must show that
\begin{align*}
\text{FINDP} \le_\pi \text{ends}\{\sigma.\; \text{leastp}(\sigma, f)\} + \text{ends}\{\sigma.\; \sigma(f) = \text{len}(\text{array})\}.
\end{align*}
where
\begin{align*}
\text{leastp}(\sigma, x) =  (\forall v<\sigma(x). \lnot P(\sigma(\text{array}[v])))\land P(\sigma(\text{array}[\sigma(x)])).
\end{align*}
The interesting part of this proof is the verification
